A boundary-integral framework to simulate \\ viscous erosion of a porous medium

### 2018-04-12 docs/methods/erosion_methods.tex
NM

% close contact, close proximity, nearby

The methods developed here will allow us to simulate the erosion of many bodies, as demonstrated in Fig.~\ref{fig:50bodies} for the case of 50 bodies (of randomly assigned initial sizes and positions). Here, the erosion of solid material arises from the mechanical stress induced by the surrounding flow. In particular, we take the local rate of erosion to be proportional to the magnitude of the shear stress---a law that has been verified by laboratory experiments~\cite{ris-moo-chi-she-zha2012}. As seen in Fig.~\ref{fig:50bodies}, erosion not only diminishes the size of the bodies, but also alters their shapes considerably. Notice, in particular, that many bodies develop corner-like features and nearly flat faces, with straight channels intervening between them. In the figure, color represents the vorticity of the surrounding flow, which reduces to shear on solid boundaries and thus provides a convenient tool for visualizing local erosion rates.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.80 \textwidth]{./figs/50bod.pdf}
\caption{\label{fig:50bodies} Simulation of 50 bodies eroding in Stokes flow under the action of fluid-mechanical shear. The flow is horizontal (left to right) and the 6 snapshots are evenly spaced in time. Color represents vorticity, which provides a convenient way to visualize local shear rates. Erosion not only diminishes the size of the bodies but also alters their shapes considerably. A few well-defined channels develop in the space between bodies. Several of the bodies vanish in finite time and the simulation continues.
}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure generated from data file 50circ512p5, 
%which uses 50circ512.in with epsfac = 30, sigfac = 10, dt = 1e-4, fixpdrop = 1

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
To perform this simulation of Stokesian erosion of a porous media, two fast and accurate methods are employed---one to solve the fluid equations and the other to time advance the eroding grains.  The fluid equations are reformulated in terms of a boundary integral equation (BIE)~\cite{poz1992}.  Since the boundary conditions are Dirichlet, we use a double-layer potential formulation so that the resulting BIE is second-kind.  This second-kind BIE is discretized with the spectrally accurate trapezoid rule~\cite{tre-wei2014} which results in a well-conditioned dense linear system for an unknown density function, and the linear system is solved using fast-multipole-accelerated GMRES~\cite{saa-sch1986}.  Once the density function is computed, the shear stress is computed with a spectrally accurate odd-even quadrature rule~\cite{sid-isr1988}.  In addition, the pressure and vorticity are computed in the fluid bulk (Fig.~\ref{fig:50bodies}) with an upsampled trapezoid rule to assist with visualizing the effect of erosion in porous media.

Next, the computed shear stress is used to evolve the solid boundaries.  Tracking the interface of fluid-structure interaction problems with a long time horizon is particularly difficult as the mesh can become distorted and tangled.  Furthermore, the process of erosion naturally leads to regions of high curvature and even corners.  The interface tracking issues are addressed in two ways: first, we introduce a tangential velocity that keeps the grid equispaced with respect to arc length but that does not affect the interface dynamics; second, we introduce a curvature-penalizing term into the normal interface velocity to suppress high-frequency oscillations associated with nearly singular curvature.  The interface is parametrized in the so-called {\thL} coordinate system so that the curvature penalization appears as a diffusive term that can easily be handled in a numerically stable fashion with an exponential integrator.

We supplement our primarily computational study with a few analytical results. In particular, we find a single body eroding in Stokes flow tends to a slender, for-aft symmetric form with sharp corners at its front and rear. By appealing to the principle that erosion more evenly distributes the shear stress along the surface, we can analytically predict the opening angle formed at the front and rear. In addition, drag laws from slender-body theory allows us to derive scaling laws for the rate of the body's area and drag reduction. These analytical results, in addition to having value in their own right, allow us to conclusively validate our simulation.

\paragraph{Limitations}
We make two fundamental design choices in our simulation: first, we restrict attention to two spatial dimensions, and second the bodies are considered immobile (i.e.~fixed to their initial locations). The restriction to two dimensions allows us to simulate the erosion of many bodies (up to $50$ in this manuscript) with high resolution, giving us the ability to accurately characterize the erosion of a porous medium. While BIEs are available in three dimensions, the extra computational cost limits the number of bodies that could be practically simulated, making porous-medium simulations less feasible. The second assumption of immobile bodies affords a simple framework to study the effects of erosion in isolation. While other effects, such as sedimentation or transport, undoubtedly play a role in real applications, these effects would introduce additional complexity, both conceptually and numerically. We therefore defer the study of these effects to future work.

\paragraph{Related Work}
In closely related work, Mitchell \& Spagnolie used a traction integral equation formulation to simulate viscous erosion of a single body in three dimensions~\cite{mit-spa2016}. The flexibility of their formulation enabled simulations of a {\em mobile} body eroding in an {\em aribtrary} background flow (e.g.~uniform flow, shear flow, etc.).  Another advantage of the traction formulation is that is directly provides the shear stress that is needed to evolve surfaces.  In the present work, we aim to simulate the erosion of a porous medium and, further, choose to fix the bodies to their initial locations. Thus, rather than a {\em single, mobile} body as considered in~\cite{mit-spa2016}, we aim to simulate the erosion of {\em many, immobile} bodies. Although the traction formulation has several attractive features, we instead employ a standard double-layer formulation since it allows us to easily calculate bulk quantities (e.g.~vorticity, pressure, and velocity) that will assist our understanding of the erosion process.

>>>>

Ultimately, the methods developed here will lead to high-fidelity simulations of multi-body Stokes erosion as illustrated in Fig.~\ref{fig:50bodies} for the case of 50 bodies (of randomly assigned initial sizes and positions). Here, the erosion of solid material arises from the mechanical stress induced by the surrounding flow. In particular, we take the local rate of erosion to be proportional to the magnitude of the shear stress---a law that has been verified by laboratory experiments~\cite{ris-moo-chi-she-zha2012}. As seen in Fig.~\ref{fig:50bodies}, erosion not only diminishes the size of the bodies, but also alters their shapes considerably. In particular, several bodies develop corner-like features and nearly flat faces, with straight channels intervening between them. In the figure, color represents the vorticity of the surrounding flow, which reduces to shear on solid boundaries and thus provides a convenient way to visualize local erosion rates. Several bodies vanish in finite time and the simulation continues without interruption.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.80 \textwidth]{./figs/50bod.pdf}
\caption{\label{fig:50bodies} Simulation of 50 bodies eroding in Stokes flow under the action of fluid-mechanical shear. The flow is horizontal (left to right) and the 6 snapshots are evenly spaced in time. Color represents vorticity, which provides a convenient way to visualize local shear rates. Erosion not only diminishes the size of the bodies but also alters their shapes considerably. A few well-defined channels develop in the space between bodies. Several of the bodies vanish in finite time and the simulation continues without interruption.
}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure generated from data file 50circ512p5, 
%which uses 50circ512.in with epsfac = 30, sigfac = 10, dt = 1e-4, fixpdrop = 1

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
To perform the above simulation, two fast and accurate methods are employed---one to solve the fluid equations and the other to evolve the eroding grains.  The fluid equations are reformulated in terms of a boundary integral equation (BIE)~\cite{poz1992}.  Since the boundary conditions are Dirichlet, we employ a double-layer potential so that the resulting BIE is second-kind.  This second-kind BIE is discretized with a spectrally accurate trapezoid rule~\cite{tre-wei2014} which results in a well-conditioned dense linear system for an unknown density function. The linear system is solved using fast-multipole-accelerated GMRES~\cite{saa-sch1986}.  Once the density function is determined, we compute the shear stress with a spectrally accurate odd-even quadrature rule~\cite{sid-isr1988}.

At the next stage, we must use this shear stress to advance solid boundaries. Tracking fluid-structure interfaces with long time horizons presents numerical challenges in that the underlying mesh can become distorted and tangled. Furthermore, the process of erosion naturally leads to regions of high curvature and even corners.  We address these tracking issues in two ways: first, we introduce a tangential velocity that keeps the grid equispaced with respect to arc length but that does not affect dynamics; second, we suppress high-frequency oscillations in the interface evolution via a curvature-penalization. The interface is parametrized in the so-called {\thL} coordinate system so that the curvature penalization appears as a diffusive term. This term can be easily be handled in a numerically stable fashion with an exponential integrator. 
Once the simulation is complete, we perform a post-processing step in which we compute bulk properties, such as vorticity and pressure, using an upsampled trapezoid rule.

We supplement our primarily computational study with a few analytical results. In particular, our simulation shows that a single eroding body tends to a slender, for-aft symmetric form characterized by nearly uniform shear stress along its surface. This body features sharp corners at its front and rear stagnation points. The principle of long-time uniform stress combined with some local analysis leads to an analytical prediction for the opening angles formed at the front and rear. Additionally, drag laws from slender-body theory allows us to derive scaling laws for the rate of the body's area and drag reduction. These analytical results, in addition to their own intrinsic value, allow us to conclusively validate the numerical simulation.

\paragraph{Limitations}
We make two fundamental design choices in our simulation: first, we restrict attention to two spatial dimensions, and second the bodies are considered immobile (i.e.~fixed to their initial locations). The restriction to two dimensions allows us to simulate the erosion of many bodies (up to $50$ in this manuscript) with high resolution, enabling accurate characterization of porous-medium erosion. While BIEs are available in three dimensions, the extra computational cost limits the number of bodies that could be practically simulated and makes porous-medium simulations less feasible. Second, the assumption of immobile bodies affords a simple framework to analyze the effects of erosion in isolation. While other mechanisms, such as sedimentation or transport, undoubtedly play a role in real applications, these effects would introduce additional complexity, both conceptually and numerically. We therefore defer to future work the incorporation of these additional mechanisms.

\paragraph{Related Work}
In a closely related study, Mitchell \& Spagnolie used a traction integral equation formulation to simulate viscous erosion of a single body in three dimensions~\cite{mit-spa2016}. The flexibility of their formulation enabled simulations of a {\em mobile} body eroding in an {\em aribtrary} background flow (e.g.~uniform flow, shear flow, etc.).  An advantage of the traction formulation is that is directly provides the shear stress that is needed to evolve surfaces.  In the present work, we aim to simulate the erosion of a porous medium and, further, choose to fix the bodies to their initial locations. Thus, rather than a {\em single, mobile} body as considered in~\cite{mit-spa2016}, we aim to simulate the erosion of {\em many, immobile} bodies. Although the traction formulation has several attractive features, we instead employ a standard double-layer formulation since it allows us to easily calculate bulk quantities (e.g.~vorticity, pressure, and velocity) that will assist our understanding of the erosion process.
